---
title: "Poisson Regression and Applications"
author: "Adam Dawson, Sarah Onstad-Hawes, and Asare Buahin"
date: "12/13/2019"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
 

##  Introduction: Main Project Questions

- What is the Poisson Regression and what are it's applications?

- Can we fit the Poisson regression to our data set and use it to make any inferences about the relationships we see?


## Introduction: Our Data Set


 - "Contact with Medical Doctors"
 
    - Cross-Sectional data collected in North Carolina between 1977-78
 
    - Examines 20186 observations of individuals
 
      - Randomly selected subset of 2000 observations for ease of computation
    
    - Collected from RAND Health Insurance Experiment (RHIE), which is the longest and largest socially controlled experiment regarding medical care (Price, D. 2002)
    
    - Our Response Variable of interest is mdu, which captures the number of times an individual visited a medical health profession during the study
    
    - Data Set also contains 14 other variables of interest
      

## Introduction: Our Research Question

- How is the number of doctor visits impacted by various factors in an individual's life?

  - Specifically, we want see how one's age, sex, income, physical limitations, and present diseases influence their ability to seek out medical care.  
  
- Use the Poisson Regression to model the data and see what relationships, if any, exist in between our variables.  


## Methodology: The Poisson Distribution

- A probability function which is especially useful for count data
    - $Y$ is a variable with discrete outcomes ($0,1,2,...$) where high counts for $Y$ are rare
    - $f(Y) = \frac{\mu^{Y}*e^{-\mu}}{Y!}$
    - $E[Y] = \mu$
    - $P(Y = y) = \frac{\mu^{y}*e^{-\mu}}{y!}$
    - $\mu$ also sometimes noted as $\lambda$
    
## Methodology: Necessary Conditions for the Poisson Distribution

- The Mean and Variance of Y are equal
  - $E[Y] = V[Y] = \mu$
  
- Independence
  - An event A occuring does not impact event B from occuring
  
- Each observation is recorder over the same fixed period of time.

- Data Set is not overloaded with zero counts. 

## Methodology: The Poisson Regression

- Poisson Regression finds estimates for the linear equation relating the log of a response variable to predictors.
  - $log(\hat{\mu}) = \beta_{0} + \beta_{1}X_{1} + \beta_{2}X_{2} + ... + \beta_{p}X_{p}$
  - $\hat{\mu} = e^{\beta_{0} + \beta_{1}X_{1} + \beta_{2}X_{2} + ... + \beta_{p}X_{p}}$
-  Allows us to see how a marginal change in our predictor variable impacts the estimated count of our variable.  
- Coefficients found using the Maximum Likelihood Estimate process

- Coefficients evalued for statisical significance using z-statistic and corresponding p-value

- We can use R glm function where we indicate the family to be "poisson" in order to obtain our fitted equation and coefficient estimates. 


## Results and Conclusion: An Introduction to our Data

- Observational Units: Individuals in North Carolina

- Overall, we have chosen five predictor variables to estimate our response variable
- Response Variable: 
    - mdu measures the number of doctors visits one person attends in a year: Count
- Predictor Variables used:
  - Linc denotes a person's yearly log(income): Quantitative
  - Age denotes a person's age at the time of the study: Quantitative
  - Physlim denotes if a person has any sort of  physical limitation: Categorical, Binary
  - ndiseases denotes the number of diagnosed diseases a person has at the time of the study:Quantitative
  - Sex denotes the sex of the patient: Categorical, Binary (Male == 1, Female == 0)

## Results and Conclusions: An Introduction to our Data continued

```{r, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(tibble)
library(here)
library(Stat2Data)
library(knitr)
library(leaps)
library(GGally)

Doctor <- read.csv(here("", "DoctorContacts.csv"))
Doctor <- as_tibble(Doctor)

set.seed(1)
Doc_sample <- Doctor[sample(1:nrow(Doctor), 2000,
   replace=FALSE),]

Doctor_Summary <- Doc_sample %>%
            	summarize(count = n(), st.dev = round(sd(mdu),2),
                      	sample_mean = round(mean(mdu),2), med = round(median(mdu), 2),
                        	min = min(mdu), max = max(mdu))
kable(Doctor_Summary)
```

- $\mu$ = 2.87, $\sigma^{2}$ = 4.56

  - Over-disperion present

***

```{r, message =  FALSE, warning= FALSE, fig.height = 2, fig.width = 3, fig.cap= "Count of an Individual's Doctors Visits in a Year"}
p <- ggplot(data = Doc_sample)

p + geom_histogram(mapping = aes(x = mdu), bins = 50 )
```


- Histogram shows a distribution skewed to the right
  - A high number of 0 counts in our data set

## Results and Conclusions: Forward Selction:Step 1
```{r, message = FALSE, warning = FALSE, echo=TRUE}
regmod1<-glm(mdu~physlim, Doc_sample, family=poisson)
regmod2<-glm(mdu~ndisease, Doc_sample, family=poisson)
regmod3<-glm(mdu~linc, Doc_sample, family=poisson)
regmod4<-glm(mdu~age, Doc_sample, family=poisson)
regmod5<-glm(mdu~sex, Doc_sample, family=poisson)
p.physlim <- summary(regmod1)$coefficients[2,4]
p.disease <- summary(regmod2)$coefficients[2,4]
p.linc <- summary(regmod3)$coefficients[2,4]
p.age <- summary(regmod4)$coefficients[2,4]
p.sex <- summary(regmod5)$coefficients[2,4]
```
## Results and Conclusions: Forward Selction:Step 1  
```{r, message = FALSE, warning = FALSE}
pvalstep1 = rbind(p.physlim, p.disease, p.linc,p.age,p.sex)
pvalstep1
```
## Results and Conclusions: Forward Selction:Step 2
```{r, message = FALSE, warning = FALSE, echo=TRUE}
regmod11<-glm(mdu~ndisease+physlim, Doc_sample, 
              family=poisson)
regmod12<-glm(mdu~ndisease+linc, Doc_sample, 
              family=poisson)
regmod13<-glm(mdu~ndisease+age, Doc_sample, 
              family=poisson)
regmod14<-glm(mdu~ndisease+sex, Doc_sample, 
              family=poisson)
p.dp <- summary(regmod11)$coefficients[3,4]
p.dli <- summary(regmod12)$coefficients[3,4]
p.da <- summary(regmod13)$coefficients[3,4]
p.ds <- summary(regmod14)$coefficients[3,4]
```
```{r, message = FALSE, warning = FALSE}
pvalstep2=rbind(p.dp, p.dli, p.da,p.ds)
pvalstep2
```
## Results and Conclusions: Forward Selction:Step 3
```{r, message = FALSE, warning = FALSE, echo=TRUE}
regmod21<-glm(mdu~ndisease+physlim+age, 
              Doc_sample, family=poisson)
regmod22<-glm(mdu~ndisease+physlim+sex, 
              Doc_sample, family=poisson)
regmod23<-glm(mdu~ndisease+physlim+linc, 
              Doc_sample, family=poisson)
p.dpa <- summary(regmod21)$coefficients[4,4]
p.dps <- summary(regmod22)$coefficients[4,4]
p.dpl <- summary(regmod23)$coefficients[4,4]
```
```{r, message = FALSE, warning = FALSE}
pvalstep3=rbind(p.dpa, p.dps, p.dpl)
pvalstep3
```
## Results and Conclusions: Forward Selction:Step 4
```{r, message = FALSE, warning = FALSE, echo=TRUE}
regmod31<-glm(mdu~ndisease+physlim+linc+sex, 
              Doc_sample, family=poisson)
regmod32<-glm(mdu~ndisease+physlim+linc+age, 
              Doc_sample, family=poisson)
p.dpls <- summary(regmod31)$coefficients[5,4]
p.dpla <- summary(regmod32)$coefficients[5,4]
```
```{r, message = FALSE, warning = FALSE}
pvalstep4=rbind(p.dpls, p.dpla)
pvalstep4
```
## Results and Conclusions: Forward Selction:Step 5
```{r, message = FALSE, warning = FALSE, echo=TRUE}
regmod41<-glm(mdu~ndisease+physlim+linc+sex+age, 
              Doc_sample, family=poisson)
p.dlspa <- summary(regmod41)$coefficients[6,4]
```
```{r, message = FALSE, warning = FALSE}
pvalstep5=rbind(p.dlspa)
pvalstep5
```

## Results and Conclusions: Summary Output of model
```{r, message = FALSE, warning = FALSE}
model <- glm( mdu ~ ndisease + physlim + linc + sex + age,  Doc_sample, family = poisson) 
summary(model)
```
## Results and Conclusions: Final Model
- From our knowledge on Poisson Distribution, we can write out the regression model like so
$log(MDU) = -1.03+0.029NDisease + 0.435PhysLim + 0.184linc-0.245Male+0.004Age$

- Which is equivalent to 
$MDU = e^{-1.03+0.029NDisease + 0.435PhysLim + 0.184linc-0.245Male+0.004Age}$

## Results and Conclusions: Another Application

- In addition to regression, we can use the poisson distribution to find probabilities of specific count occurences or ranges of count occurences in the data.

```{r, echo = TRUE}
Probof4ormore = ppois(4,lambda=2.86, lower.tail=FALSE)
Probof4ormore
```

- This gives us the probability that a person will go to the doctor's office 4 or more times in a year. We can edit the lower.tail component of this code to give us the probablity that a person will go to the doctor's office 4 or less times in a year


## Discussion and Critiques: Model Assumptions

- We have good reason to question a few necessary assumptions of the Poisson Distribution in regards to our data set

- Mean and Variance are not equal $E[Y] = 2.8$ and $V[Y] = 4.5$
  - Over-dispersion, which could indicate that we should use a different model

- We have a large number of observations for which the count is 0

- Consider using a different model to fit to our data
  - Zero-Inflated Poisson Regression
    - accounts for large number of 0 counts in data set
  - Zero-Inflated Negative Binomial Regression
    - helps with both larger number of 0s and over-dispersion


## Discussion and Critiques: Inference

- With our data being so significant (perhaps abnormally), we were unsure of our ability to use the fitted equation with any confidence

- With a high number of possible predictors, we could have ommitted some very significant variables from the regression, which could be vital in understanding what impacts utilization of healthcare.

- Data set is from the 1970s
  - Comprehensive, but outdated
  - If using data to enact policy shifts or structural changes, our data analysis might not be "in touch" enough with the current climate of health care. 
 
 - Data is collected from N.C. only, which limits the scope in which we can make inferences.
  - Perhaps some cultural/societal impacts in N.C. region impacted our data
  - US healthcare system unique, so inferences at most apply only to US.
  
 
## Conclusion

- Poisson Regression is useful for analyzing count data, but it is crucial to check the data set and see that conditions are met prior to analysis and inference

- Having a large sample size is helpful, but also must be aware of how that potentially impacts analysis.

- Working on real world data is tough and messy. Takes a lot more time and careful thought
  - No exact, clear path for analysis
- "One step forward, two steps back"

## References

https://stats.idre.ucla.edu/r/dae/poisson-regression/
https://stats.idre.ucla.edu/stata/output/poisson-regression/
https://stats.idre.ucla.edu/r/dae/zip/
https://www.sciencedirect.com/science/article/pii/S0167629602000085
https://www.dataquest.io/blog/tutorial-poisson-regression-in-r/#:~:targetText=Poisson%20Regression%20models%20are%20best,where%20the%20outcomes%20are%20counts.&targetText=Poisson%20Regression%20helps%20us%20analyze,the%20count%20or%20a%20rate).






